\Large \bf Cyclone: Replication Middleware for NVM Applications

### 2016-09-23 paper/paper.tex
Tweak graph placement.
Submission related updates.

\subsection*{Abstract}
The introduction of non-volatile directly attached memory in datacenter - in
the form of NVDIMMs and new memory technologies allows efficient data structures
to be built on non-volatile heaps that provide good performance coupled with
durability. However high availability remains a problem for these primarily
single machine solutions. While transactional interfaces built over RDMA
capable networking hardware offer a viable alternative, they are often
unavailable in datacenters. Cyclone solves this problem by providing transparent
replication middleware that to be used with Intel's Non Volatile Memory Library 
(NVML) to construct durable data-structures that map keys to values.
Cyclone is designed around an optimized userspace networking stack to reduce
latency and can scale its consensus protocol across cores and network ports to
provide good throughput under load. We show that Cyclone can replicate as many
as 8 million updates a second using just 4 commodity 10 GigE ports on a machine.

\section{Introduction}
Memory technology in datacenters is due to undergo a paradigm shift with the
increased use of directly addressable non-volatile memory both in the form of
battery backed non-volatile DIMMs~\cite{farm} as well as newer memory types
such as 3D XPoint~\cite{pmfs, bpfs}. In anticipation, libraries that allow
programmers to build durable data structures on non-volatile heaps have become
available~\cite{mnemosyne, nvheaps, cdds}  - an example of a such a
library that we have used in this paper is Intel's non volatile memory library
(NVML)~\cite{nvml}. A missing piece however is a component to make such durable 
data structures highly available using replication across a commodity network
such as ethernet. Existing solutions that offer good performance are biased
towards pure transactions across RDMA with a dependence on external services
(such as zookeeper~\cite{zookeeper}) and application specific code for
fault recovery~\cite{farm, htm}.
This is a problem for programmers using NVML to construct durable and available 
data structures where one cannot apriori guarantee the availability of
RDMA (for e.g., when running on externally hosted environments or public
clouds).

Cyclone is a prototype replication middleware that aims to demonstrate that one
can add fault tolerance transparently to a specific class of client server
applications written on top of libraries such as NVML. We specifically target
key-value maps in this paper and in that context make two key
contributions. First, we propose a simple API that enables server state to be
replicated without the programmer needing to worry about fault recovery. This
allows one to start by building a concurrent durable data structure using
libraries such as NVML and add availability at a later stage using Cyclone. The
second contribution is to show that one can achieve reasonably good replication 
performance using commodity ethernet via innovations in the networking stack and
by scaling replication with the application. This result should be a useful
design point for developers looking to deploy applications on non-volatile heaps
in commodity ethernet-based environments such as public clouds.

\section{Single Node Programming Model}
Cyclone assumes a concurrent durable application on non-volatile heap as the
starting point. A number of libraries and filesystems make this possible
today~\cite{nvml, dax, pmfs, mnemosyne, nvheaps, cdds}. Cyclone currently
supports applications built on top of Intel's Non Volatile Memory Library
(NVML~\cite{nvml}). We describe NVML's programming model in this section to
provide adequate background for the rest of this paper. NVML requires 
that code that makes changes to non-volatile heap is enclosed in a crash
consistent transaction such that either all of the transaction executes or none
of it executes in the event of power failures.
%This is critical to maintaining consistency as the data structure, being
%durable, must survive across power failures.
Figure~\ref{fig:example} illustrates the use of NVML to build a persistent
linked list. Updates to the linked list are enclosed in a NVML transaction to
avoid corruption due to power failure. It is also worth noting that NVM
libraries such as NVML usually allow remapping non-volatile heaps to a different
virtual memory address range across restarts (e.g., using {\tt TOID} macros to
implement fat pointers in NVML).

@@

In order to avoid an additional copy of the incoming RPC call contents from DRAM
to persistent memory, Cyclone allocates space for packet buffers on non-volatile
heap. However, since packet data is cached, Cyclone ensures durability
(or persistence) of this data by explicitly flushing the dirty
cachelines~\cite{pmfs}. Cyclone maintains its log as a circular sequence of
pointers to packet buffers and therefore the append operation is completed by
adding a pointer to the received packet and flushing the cacheline containing
the pointer.

We exploit the fact that DPDK allows prepending data to a packet
without physically moving the data in a packet buffer. The received packet
contains only the RPC call from the client. We prepend a RAFT related header
(containing the previous log index, previous log term, leader's term and
leader's commit point~\cite{raft}) to packet buffer itself. To this RAFT header 
we then prepend an IPv4 header. We assign pre-determined values to the
IP address fields in order to exploit 5-tuple steering at the receiving NIC for 
the placement of the incoming packet in the right queue.
This composite packet buffer is then sent \emph{unchanged} to every replica.
To properly route the packet through switches, we allocate a small buffer
containing an ethernet header for each follower replica and \emph{chain} the
buffer in front of the prepared packet - thereby ensuring that we do not need
to change the contents of the packet itself for every replica.

@@

\section{A Durable Replicated Hash Table}
\label{sec:example}
We now demonstrate how Cyclone can be used in practice by putting together an
example of a linearizable hash table - a useful and common data structure.
[memcached]. The baseline is a durable hash table written in NVML that runs on a
single machine in client-server mode. The server side maintains the durable
open-addressed hash table and services client RPC calls to insert and delete
key-value pairs and lookup keys. Such a hash table is simple to derive from a
volatile hash table and readily available in NVML [nvml.io] as an example [src
  code link]. To use that (non client-server) example with Cyclone we added a
simple RPC call format, manually marshalled and demarshalled the few
arguments to the RPC call in an RPC call handler we wrote - a few tens of lines
of code. Finally, we added a simple client driver program (a few tens of
additional lines of code).

@@

We support distributed transactions in our hash table by allowing the user to
specify a group of insert/delete/lookup operations that must occur
atomically. We use the existing bucket locks to implement a simple two phase
locking protocol to implement the transction. A client makes an RPC call to the
server specifying the transaction and the server executes it on the client's
behalf. Execution of the transaction has four distinct phases: acquire locks,
verify values, apply updates and finally release locks.

We guard against faults by executing the RPC call on a quorum of machines via
Cyclone. Since execution of the RPC call leaves \emph{externally visible side
  effects} by virtue of the RPC calls it makes we require output
commit~\cite{output_commit}. This is achieved if the execution is guaranteed to
complete once it starts. This is simply achieved by a flag in the RPC call that
turns off decoupled execution (Section~\ref{sec:decouple}). The flag causes
execution to start only after the RPC call has been successfully replicated.

@@

\section{Evaluation}
We evaluate Cyclone against its design goals by using it for up to 3-way
replication in a cluster of twelve 48 core Xeon servers. Three servers run
Cyclone's server side code, while the remaining machines run multithreaded
clients to load the system for our studies. The server side machines are each
equipped with four commodity 82599 10 Gbit ethernet controllers, while each
client machine has one 82599 ethernet controller. All the machines are connected
through the same commodity 10Gbit switch. We use 8 cores on the server machines
for running 8 instance of RAFT, leaving 39 cores to run application code. The
last remaining core is reserved for management code. We use NVML with the non
volatile pool held in DRAM, which proxies for the performance we would expect
from NVDIMMs. We evaluate Cyclone in three stages.

First, we evaluate pure replication performance by replicating RPC calls. There
is crash consistent transaction that remembers the RPC call sequence number -
every RPC call is accepted and once replicated a response is sent to the
client. Figure~\ref{fig:pure_rep} shows the results. We consider two cases, one
where the clients send an RPC call with no payload and one where they add a 512
byte dummy payload. For the no payload case the bottleneck is the message
handling rate. We note that Cyclone can handle small message replication rates
as high as \emph{8 million messages a second}. In the case of RPC calls with 512
byte payloads, the bottleneck is the available network bandwidth. We plot
transmit throughput derived from the message rate as the rate at which data is
sent to replicas from all the 8 quorum leaders and back to the clients.
It is interesting to note that with 3 replicas Cyclone achieves \emph{more than
  40 Gbps} of bandwidth. The reason for this is that the quorum leaders can end
up residing on different machines. Although replication traffic is still limited
by the receive bandwidth the responses to the clients, originating from
different machines have access to independent 4x10GigE NICs, allowing total
transmit throughput to exceed 40 Gbps. Although not discussed in this paper,
Cyclone's control plane is designed to exploit this by trying to spread quorum
leaders across machines.

@@

\begin{figure}
\begin{tabular}{cc}
\begin{minipage}{0.25\textwidth}
  \includegraphics[width=\textwidth,height=3cm]{results/replication_mops_0.pdf}
\end{minipage}&
\begin{minipage}{0.25\textwidth}
  \hspace{-0.2in}
  \includegraphics[width=\textwidth,height=3cm]{results/replication_gbps_0.pdf}
\end{minipage}\\
\begin{minipage}{0.25\textwidth}
  \includegraphics[width=\textwidth,height=3cm]{results/replication_mops_512.pdf}
\end{minipage}&
\begin{minipage}{0.25\textwidth}
  \hspace{-0.2in}
  \includegraphics[width=\textwidth,height=3cm]{results/replication_gbps_512.pdf}
\end{minipage}
\end{tabular}
\caption{Pure Replication}
\label{fig:pure_rep}
\end{figure}

Next, we consider the effect of adding NVMLs crash consistent
transactions. Figure~\ref{fig:cc_rep} shows the effect of adding a crash
consistent transaction where the application still does nothing but Cyclone
remembers the last sequence number and RPC call result (the application simply
returns the received payload). Throughput is now bottlenecked by NVML allocator
performance at just under a million messages a second. The final result is
adding the hash table in the picture with the clients updating the values of
keys already present in the hash table - a stable workload for easy
comparison. As Figure~\ref{fig:app_rep} shows performance stays at the just
under 1M calls a second mark for pure updates. For comparison we also show a
pure read only case where the clients only query the value of a key - there is
no replication \emph{nor} any crash consistent transaction applied. It is
interesting to note that the pure read only case is bottlenecked at the same
point as the pure replication case of 8 million operations a second.

These results lead us to conclude that Cyclone's provides adequate performance
headroom for replication to NVML applications. In fact our current focus in the
light of these results is towards improving application performance on top of
NVML. NVML is a general purpose library written for a variety of substrates
including traditional SSD based storage and there is adequate room for
improvement of performance for applications running on directly attached
persistent memory with close to DRAM performance.

\begin{figure}
\begin{tabular}{cc}
\begin{minipage}{0.25\textwidth}
  \includegraphics[width=\textwidth,height=3cm]{results/cc_mops.pdf}
  \caption{Crash Consistency + Replication}
  \label{fig:cc_rep}
\end{minipage} &
\begin{minipage}{0.25\textwidth}
  \includegraphics[width=\textwidth,height=3cm]{results/app_mops.pdf}
  \caption{Crash Consistent Replicated Hash Table}
  \label{fig:app_rep}
\end{minipage}
\end{tabular}
\end{figure}

\section{Conclusion}
Cyclone allows developers to add availability to applications built on top of
libraries such as NVML and achieve good performance on commodity networking
hardware. A key question is whether the ideas in Cyclone extend to more than
just key-value maps - perhaps to application such as filesystems ? We believe
that the requirement for scaling across multiple replication quorums can be
satisfied by applications where the scalable commutativity
rule~\cite{scalable_commutativity} applies to common case primitives. Although
our intitial release of Cyclone will be targeted towards key value style
applications, our future research is targeted towards investigating this
hypothesis.
\newcommand\myurl[2]{\url{#1}}
\bibliographystyle{acm}
\bibliography{paper}

>>>>

\subsection*{Abstract}
The introduction of non-volatile directly attached memory in datacenters - in
the form of NVDIMMs and new memory technologies allows efficient data structures
to be built on non-volatile heaps that provide good performance coupled with
durability. However high availability remains a problem for these primarily
single machine solutions. While transactional interfaces built over RDMA capable
networking hardware offer a viable alternative, such hardware is often
unavailable in datacenters. Cyclone offers a solution to this problem by
providing transparent replication middleware that can be used with Intel's Non
Volatile Memory Library (NVML) to construct durable data-structures that map
keys to values.  Cyclone is designed around an optimized userspace networking
stack to reduce latency and can scale its consensus protocol across cores and
network ports to provide good throughput under load. We show that Cyclone can
replicate as many as 8 million updates a second using just 4 commodity 10 GigE
ports on a machine.

\section{Introduction}
Memory technology in datacenters is due to undergo a paradigm shift with the
increased use of directly addressable non-volatile memory both in the form of
battery backed non-volatile DIMMs~\cite{farm} as well as newer memory types
such as 3D XPoint~\cite{pmfs, bpfs}. In anticipation, libraries that allow
programmers to build durable data structures on non-volatile heaps have become
available~\cite{mnemosyne, nvheaps, cdds}  - an example of such a
library that we have used in this paper is Intel's non volatile memory library
(NVML)~\cite{nvml}. A missing piece however is a component to make such durable 
data structures highly available using replication across a commodity network
such as ethernet. Existing solutions that offer good performance are biased
towards pure transactions across RDMA with a dependence on external services
(such as zookeeper~\cite{zookeeper}) and application specific code for
fault recovery~\cite{farm, htm}.
This is a problem for programmers using NVML to construct durable and available 
data structures where one cannot apriori guarantee the availability of
RDMA (for e.g., when running on externally hosted environments or public
clouds).

Cyclone is prototype replication middleware that aims to demonstrate that one
can add fault tolerance transparently to a specific class of client server
applications written on top of libraries such as NVML. We specifically target
key-value maps in this paper and in that context make two key
contributions. First, we propose a simple API that enables server state to be
replicated without the programmer needing to worry about fault recovery. This
allows one to start by building a concurrent durable data structure using
libraries such as NVML and add availability at a later stage using Cyclone. The
second contribution is to show that one can achieve reasonably good replication 
performance using commodity ethernet via innovations in the networking stack and
by scaling replication with the application. This result should be a useful
design point for developers looking to deploy applications on non-volatile heaps
in commodity ethernet-based environments such as public clouds.

\section{Single Node Programming Model}
Cyclone assumes a concurrent durable application on a non-volatile heap as the
starting point. A number of libraries and filesystems make this possible
today~\cite{nvml, dax, pmfs, mnemosyne, nvheaps, cdds}. Cyclone currently
supports applications built on top of Intel's Non Volatile Memory Library
(NVML~\cite{nvml}). We describe NVML's programming model in this section to
provide adequate background for the rest of this paper. NVML requires 
that code that makes changes to non-volatile heap is enclosed in a crash
consistent transaction such that either all of the transaction executes or none
of it executes in the event of power failures.
%This is critical to maintaining consistency as the data structure, being
%durable, must survive across power failures.
Figure~\ref{fig:example} illustrates the use of NVML to build a persistent
linked list. Updates to the linked list are enclosed in a NVML transaction to
avoid corruption due to power failure. It is also worth noting that NVM
libraries such as NVML usually allow remapping non-volatile heaps to a different
virtual memory address range across restarts (e.g., using {\tt TOID} macros to
implement fat pointers in NVML).

@@

In order to avoid an additional copy of the incoming RPC call contents from DRAM
to persistent memory, Cyclone allocates space for packet buffers on the non-volatile
heap. However, since packet data is cached, Cyclone ensures durability
(or persistence) of this data by explicitly flushing the dirty
cachelines~\cite{pmfs}. Cyclone maintains its log as a circular sequence of
pointers to packet buffers and therefore the append operation is completed by
adding a pointer to the received packet and flushing the cacheline containing
the pointer.

We exploit the fact that DPDK allows prepending data to a packet
without physically moving the data in a packet buffer. The received packet
contains only the RPC call from the client. We prepend a RAFT related header
(containing the previous log index, previous log term, leader's term and
leader's commit point~\cite{raft}) to the packet buffer itself. To this RAFT header 
we then prepend an IPv4 header. We assign pre-determined values to the
IP address fields in order to exploit 5-tuple steering at the receiving NIC for 
the placement of the incoming packet in the correct queue.
This composite packet buffer is then sent \emph{unchanged} to every replica.
To properly route the packet through switches, we allocate a small buffer
containing an ethernet header for each follower replica and \emph{chain} the
buffer in front of the prepared packet - thereby ensuring that we do not need
to change the contents of the packet itself for every replica.

@@

\section{A Durable Replicated Hash Table}
\label{sec:example}
We now demonstrate how Cyclone can be used in practice by putting together an
example of a linearizable hash table - a useful and common data structure.
The baseline is a durable hash table written in NVML that runs on a
single machine in client-server mode. The server side maintains the durable
open-addressed hash table and services client RPC calls to insert and delete
key-value pairs and lookup keys.

A durable hash table is simple to derive from a volatile hash table and readily
available in NVML as an example~\cite{nvml_hash}. To use that (non client-server)
example with Cyclone we added a simple RPC call format, manually marshalled and
demarshalled the few arguments to the RPC call in an RPC call handler we wrote -
a few tens of lines of code. Finally, we added a simple client driver program (a
few tens of additional lines of code).

@@

We support distributed transactions in our hash table by allowing the user to
specify a group of insert/delete/lookup operations that must occur
atomically. We use the existing bucket locks to support a simple two phase
locking protocol to implement the transction. A client makes an RPC call to the
server specifying the transaction and the server executes it on the client's
behalf. Execution of the transaction has four distinct phases: acquire locks,
verify values, apply updates and finally release locks.

We guard against faults by executing the the entire transaction on a quorum of
machines via Cyclone. All machines in the quorum execute the transaction, making
calls to other quorums to execute steps in the two phase locking protocol. Since
execution of the RPC call leaves \emph{externally visible side effects} by
virtue of the RPC calls it makes we require output
commit~\cite{output_commit}. This is achieved if the execution is guaranteed to
complete once it starts. This is simply achieved by a flag in the RPC call that
turns off decoupled execution (Section~\ref{sec:decouple}). The flag causes
execution to start only after the RPC call has been successfully replicated.

@@

\section{Evaluation}
We evaluate Cyclone against its design goals by using it for up to 3-way
replication in a cluster of twelve 48 core Xeon servers. Three servers run
Cyclone's server side code, while the remaining machines run multithreaded
clients to load the system for our studies. The server side machines are each
equipped with four commodity 82599-based 10 Gbit ethernet ports, while each
client machine has one 82599-based ethernet port. All the machines are connected
through the same commodity 10 Gbit switch. We use 8 cores on the server machines
for running 8 instance of RAFT, leaving 39 cores to run application code. The
last remaining core is reserved for management code. We use NVML with the non
volatile pool held in DRAM, which proxies for the performance we would expect
from NVDIMMs. We evaluate Cyclone in three stages.

First, we evaluate pure replication performance by replicating RPC
calls. Cyclone simply replicates the contents of the RPC call without executing
it and returns a response to the client when replication is complete.
Figure~\ref{fig:pure_rep} shows the results. We consider two cases, one where
the clients send an RPC call with no payload and one where they add a 512 byte
dummy payload. For the no payload case the bottleneck is the message handling
rate. We note that Cyclone can handle small message replication rates as high as
\emph{8 million messages a second}. In the case of RPC calls with 512 byte
payloads, the bottleneck is the available network bandwidth. To show this, we
also plot transmit throughput derived from the message by adding the rate at
which data is sent to replicas from all the 8 quorum leaders to the rate at
which responses are sent back to the clients.  It is interesting to note that
with 3 replicas Cyclone achieves \emph{more than 40 Gbps} of bandwidth. The
reason for this is that the quorum leaders can end up residing on different
machines. Although replication traffic is still limited by the receive bandwidth
the responses to the clients, originating from different machines have access to
independent 4x10GigE ports, allowing total transmit throughput to exceed 40
Gbps. Although not discussed in this paper, Cyclone's control plane is designed
to exploit this by trying to spread quorum leaders across machines.

@@

\begin{figure}
\begin{tabular}{cc}
\begin{minipage}{0.25\textwidth}
  \hspace{-0.17in}
  \includegraphics[width=\textwidth,height=3cm]{results/replication_mops_0.pdf}
\end{minipage}&
\begin{minipage}{0.25\textwidth}
  \hspace{-0.37in}
  \includegraphics[width=\textwidth,height=3cm]{results/replication_gbps_0.pdf}
\end{minipage}\\
\begin{minipage}{0.25\textwidth}
  \hspace{-0.17in}
  \includegraphics[width=\textwidth,height=3cm]{results/replication_mops_512.pdf}
\end{minipage}&
\begin{minipage}{0.25\textwidth}
  \hspace{-0.37in}
  \includegraphics[width=\textwidth,height=3cm]{results/replication_gbps_512.pdf}
\end{minipage}
\end{tabular}
\caption{Pure Replication}
\label{fig:pure_rep}
\end{figure}

Next, we consider the effect of adding NVMLs crash consistent
transactions. Figure~\ref{fig:cc_rep} shows the effect of adding a crash
consistent transaction where the application still does nothing but Cyclone
durably remembers the last sequence number and RPC call result (the application
simply returns the received payload). Throughput is now bottlenecked by NVML
allocator performance at just under a million messages a second. The final
result is adding the hash table in the picture with the clients updating the
values of keys already present in the hash table - a stable workload for easy
comparison. As Figure~\ref{fig:app_rep} shows performance stays at the just
under 1M calls a second mark for pure updates. For comparison we also show a
pure read only case where the clients only query the value of a key - there is
no replication \emph{nor} any crash consistent transaction applied. It is
interesting to note that the pure read only case is bottlenecked at the same
point as the pure replication case of 8 million operations a second.

These results lead us to conclude that Cyclone provides adequate performance
headroom for transparent replication in NVML applications. In fact our current
focus in the light of these results is towards improving application performance
on top of NVML. NVML is a general purpose library written for a variety of
substrates including traditional SSD based storage and there is adequate room
for improvement of performance for applications running on directly attached
persistent memory with close to DRAM performance.

\begin{figure}
\begin{tabular}{cc}
\begin{minipage}{0.25\textwidth}
  \hspace{-0.17in}
  \includegraphics[width=\textwidth,height=3cm]{results/cc_mops.pdf}
  \caption{Crash Consistency}
  \label{fig:cc_rep}
\end{minipage} &
  \hspace{-0.37in}
\begin{minipage}{0.25\textwidth}
  \includegraphics[width=\textwidth,height=3cm]{results/app_mops.pdf}
  \caption{Hash Table}
  \label{fig:app_rep}
\end{minipage}
\end{tabular}
\end{figure}

\section{Conclusion}
Cyclone allows developers to add availability to applications built on top of
libraries such as NVML and achieve good performance on commodity networking
hardware. A key question is whether Cyclone can be used for other applications
such as filesystems ? We believe that the requirement for scaling across
multiple replication quorums can be satisfied by applications where the scalable
commutativity rule~\cite{scalable_commutativity} applies to common case
primitives. Although our intitial release of Cyclone will be for
key value style applications, our future research is based on this hypothesis.
\newcommand\myurl[2]{\url{#1}}
\bibliographystyle{acm}
\bibliography{paper}
